Omics-based Single Step Trait Prediction

### 2017-09-07 paper/manuscript.tex
paper, update discussion to scenario a

\Genetics2level{Superiority of single step models}
\subsection{Inbred lines}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 62$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to the 211 tropical/subtropical lines and excluded 
all other genotypes from our analyses whereas \citeNP{Guo2016} used genotypes 
from four different subpopulations, modeled as a fixed effect, and used only a 
subset of 157 genotypes from the set of tropical/subtropical lines.
2) We applied quality checks for the predictor data after generating the subset
of 211 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Maize hybrids}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least of a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
Evaluations of core sets for the parental inbred lines of the maize hybrids were 
outside the scope of this study, but results on the inbred lines have shown that 
homogeneous coverage of the genetic space by the incomplete predictor can 
considerably leverage the predictive ability in single step prediction compared 
to using only the complete predictor.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
When using core sets of varying sizes, which were established with the
constraint of maximizing the average genetic distance among their members, we
noticed strong fluctuations in predictive abilities when moving from small core
sets to large core sets.
For traits that were better predicted by transcriptomic than by genomic
information in the core set, predictive abilities typically peaked around
core set size of 0.5/0.6, then dropped sharply at 0.7 and started to increase
again with a size of 0.9.
A principal component analysis showed that, at a core set size of 0.7, 
genotypes assigned to the A3 ancestral population (Fig.~\ref{fig:PopStructure})
were largely elements of the core set whereas genotypes with a large share at 
the A1 ancestral population were predominantly excluded from the core set.
At a core set size of 90\% of the original size, the majority of the A1
genotypes were finally covered by the core set as well.
This result suggests a negative bias in the imputation of genotypes from the A1
population and could explain the drop in predictive ability at a core set size
of 0.7, which is in agreement with the reduced predictive abilities observed
when stripping A1 genotypes from their transcriptomic information
(Fig.~\ref{fig:InbredResults}B).

\section*{Conclusions}
We have successfully applied single step prediction to a hybrid plant data set 
while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could show the importance of homogeneous coverage
of the genetic space by genotypes covered by both predictors on predictive
ability.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.

>>>>

\Genetics2level{Superiority of single step models}
\subsection{Exp2}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 54$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp1}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in \textit{Exp2} we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20--30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.

\section*{Conclusions}
We have successfully applied single step prediction to a hybrid plant data set 
while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could elucidate the influence of differences
between two predictors on predictive ability when used in a single step
approach.
When the discrepancy in predictive ability between the complete and the
incomplete predictor was skewed profoundly in favor of the incomplete predictor,
such as for pedigree and transcriptomic data, single step prediction was shown
be highly successful.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.
